The Evolution of Phenotypically Invariant Genetic Networks

### 2017-06-16 handnotes.tex
minor edits
added words to examples, edited wrt to plr notes, changed plots.

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage[super]{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

\begin{multicols}{2}

\section*{Abstract}
  \textbf{
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function. 
}
\section*{Introduction}

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice. Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

@@

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

In this paper we propose a framework to study the evolution of biological systems. To begin, we focus on the evolution of an idealized population. We consider the evolution of a perfectly adapted, large population, evolving in a static environment for an infinite number of generations. Under these ideal circumstances, we expect to observe a ``conservation of phenotype,'' where the population explores the manifold of phenotypically-invariant (or symmetric) genetic and developmental architecutres. We would like to understand which parameters influence the distribution of a population along the manifold of phenotypically invariant genetic systems. Further, we can show how dispersion along this manifold contributes to speciation and evolvability.

\section*{The Model I: Gene Networks as Linear Systems}

  Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. Thus we say that some function $\phi(t)$ is a phenotype where, 
  \begin{align}
    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
  \end{align}
  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.

@@

  We describe the impulse response as, 
  \begin{align}
    h(t) = C e^{A t} B
  \end{align}
  where $A$ is a gene network -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.

  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. However, for simplicity in exposition (and without loss of generality) we set $\ell = 1$, so that $B$ and $C$ are simply vectors of length $n$.

@@

  \begin{example}[Oscillating Gene Network: Circadian Rhythm]
    \begin{align*}
      A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
    \end{align*}
    \begin{align*}
      \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
        0 & 1 \\ 
       -1 & 0 
        \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
          \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
     \end{align*}

      \begin{align*}
        h(t) = \sin(t) + \cos(t)
      \end{align*}

    \begin{figure}[H]
        \begin{center}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

@@

            \end{tikzpicture}
        \end{center}
        \caption{
            Diagram of Example \ref{ex:osc} in the text.
            \plr{explain what arrows mean if nec}
            \label{fig:oscillator_diagram}}
    \end{figure}
    \begin{figure}[H]
      \centering
      \includegraphics[width=0.5\textwidth]{osc_impulse}
    \end{figure}
  \end{example}

@@

  Systems with identical external dynamics do not necessarily have identical internal dynamics. Any linear and minimal system -- minimal, informally meaning that the system's external dynamics are achieved with the fewest possible number of internal components -- has identical external dynamics up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C e^{A t} B
  \end{align}
  Two systems, $\Sigma = \{ A, B, C \}$, and $\bar{\Sigma} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same dynamics if they are related by a change of coordinates.

@@

  \begin{example}[External Equivalence does not imply internal equivalence]
    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
  \end{figure}
  \end{example}

@@

@@

  \plr{NOT TRUE?:}
  However, the process always has the property that
  changes in $A$ accumulate at constant (mean squared) rate:
  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
  \begin{aligned}
      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 t + O(t^2) .
  \end{aligned}

  \begin{align}
    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
  \end{align}
  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$. 
\jss{Change in $\tau$ is tracking movement of population mean, not ``macro'' evolutionary change, as interpreted previously.}
  \begin{align}
    \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
    A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
  \end{align}
  \begin{example}[All Phenotypically Equivalent Oscillators]
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix} \\
        B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
          h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\end{tikzpicture}
    \end{figure}
  \end{example}
  \section*{Phenotypic Invariance}
    First we focus on a simple evolutionary scenario: a large population, perfectly adapted, and in a constant environment. In this circumstance we expect phenotype to be conserved throughout evolutionary time. As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures. These phenotypic variations should yield distinct signatures. Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$. Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.

    Presently we ask two questions, (1) holding $\phi(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?

@@

     \begin{align}
        C \lor B &= I \\
        IV^{-1} &= I \\
        \iff V &= I
      \end{align}
      \end{example}
      \begin{example}[All Non-Minimal Gene Networks can Drift]

@@

      A diploid organism's gene network is simply the average of both of its gene network copies; one from each parent. Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes randomly. Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes. The second generation hybrids (F2s), however, will be the product of a meiosis between parental haplotypes followed by an averaging of gametes.

@@

      %\begin{align*}
       % G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
      %\end{align*}
      and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i, \tau, \hat{\tau} )$ and $G(j, \tau, \hat{\tau})$. Where each $G(\cdot)$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
      \begin{align}
        F_{2}(i, j) = \frac{G (i, \tau, \hat{\tau} ) + G( j, \tau, \hat{\tau} )}{2}
      \end{align}

@@

      The fitness of an organism can be computed by comparing its impulse response with the optimal response,
      \plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
      \begin{align}
        \mathcal{F}\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
      \end{align}
      Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.

      \begin{example}[Quantitative Biological Species Concept]
        Two populations are said to be different species if they produce low fitness offspring.

        F1s created by crossing externally equivlaent oscillators $A(0)$ and $A(2)$ have an $h_{F_{1}}(t) = e^{t}$, in contrast to both parents with $h_{F_{0}}(t) = \sin(t) + \cos(t)$. The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.

        \begin{align*}
          \mathcal{F} \left(h_{F_{0}}\right) &= 1 \\
          \mathcal{F}\left(h_{F_{1}}\right) &= 0
        \end{align*}
      \end{example}

>>>>

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

%\begin{multicols}{2}

\begin{abstract}
 % \textbf{
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function. 
%}
\end{abstract}
\section*{Introduction}

  \jss{add stuff on Haldane's Rule and Hybrid Vigor; note: F1s diverge to the 4th power, and F2s quadratically}

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice, as models allow the development of concrete numerical predictions. Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

@@

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

  In this paper we propose a framework to study the evolution of biological systems. We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution. To begin, we focus on the evolution of an idealized population. We consider the neutral evolution of a large population, evolving in a static environment. Under these circumstances we expect the population explores the manifold of phenotypically invariant (or symmetric) gene regulatory networks.

  Presently we ask two questions, (1) holding the phenotype and environment constant for evolutionary time, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?

\section*{Gene Networks as Linear Dynamical Systems}

  Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. Thus we say that some function $\phi(t)$ is a phenotype where, 
  \begin{align}
    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
  \end{align}
  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity.

@@

  We describe the impulse response as, 
  \begin{align}
    h(t) = C e^{A t} B
  \end{align}
  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.

  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$.

@@

  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}

    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}. Consider a simplified model of oscillating gene transcription. In the present framework periodic expression requires at minimum two interacting genes.

    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$. Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network). Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determiine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. (Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivlanet to its impulse response

      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

@@

            \end{tikzpicture} &
       % \end{center}
      %  \caption{
      %      Diagram of Example \ref{ex:osc} in the text.
      %      \plr{explain what arrows mean if nec}
      %      \label{fig:oscillator_diagram}}
   % \end{figure}
   % \begin{figure}[H]
    %  \centering
      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
   \end{tabular}
  %  \end{center}
      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
    \end{figure}
      We return to the evolution of such a system below.
  \end{example}

@@

  Systems with identical external dynamics do not necessarily have identical internal dynamics. Any linear and minimal system -- minimal, informally meaning that the system's external dynamics are achieved with the fewest possible number of internal components -- has identical external dynamics up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
  Two systems, $\Sigma = \{ A, B, C \}$, and $\bar{\Sigma} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same dynamics if they are related by a change of coordinates.

@@

  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks]

    Let
    \begin{align*}
      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
    \end{align*}
    with $V(\tau)$ preserving both $B$ and $C$, then
    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any valye}
    \end{figure}
 % \end{example}

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
  \end{figure}
  \end{example}

@@

First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment. In this circumstance we expect phenotype to be conserved throughout evolutionary time. As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures. These phenotypic variations should yield distinct signatures. Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$. Genetic drift registers as an increase in the intrapopulation variation in $h(t)$. 
 \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}

@@

  \plr{NOT TRUE?:}
  However, the process always has the property that
  changes in $A$ accumulate at constant (mean squared) rate:
  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
  \begin{align}
      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 t + O(t^2) .
  \end{align}

  \begin{align}
    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
  \end{align}
  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$. 
\jss{Change in $\tau$ is tracking movement of population mean, not ``macro'' evolutionary change, as interpreted previously.}
  \begin{align}
    \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
    A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
  \end{align}
 % \section*{Phenotypic Invariance}

@@

   %  \begin{align}
   %     C  = I \lor B = I \\
   %     IV^{-1} &= I \\
   %     \iff V &= I
   %   \end{align}
      \end{example}
      \begin{example}[All Non-Minimal Gene Networks can Drift]

@@

     An F1's genetic regulation is a consequence of both of its genomes, where genes and regulatory sequences from both parents are equally present. Thus we say that a diploid organism's gene network is simply the average of both of its gene network copies; one from each parent. Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes (specifically their \emph{cis}-regulatory modules) randomly. Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes. The second generation hybrids (F2s), however, will be the product of a meiosis between parental haplotypes followed by an averaging of gametes.

@@

      %\begin{align*}
       % G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
      %\end{align*}
      and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i)$ and $G(j)$. Where each $G$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
      \begin{align}
        F_{2}(i, j) = \frac{G (i) + G(j)}{2}
      \end{align}

@@

      The fitness of an organism can be computed by comparing its impulse response with the optimal response,
      \plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
      \begin{align}
        \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert dt \right\}  .
      \end{align}
      Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.

      \begin{definition}[Reproductive Incompatibility]
        According to Mayr's \emph{Biological Species Concept}, two populations can be considered different species if they are reproductively isolated, meaning crosses between them produce low fitness offspring.

        We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
        \begin{align*}
          \mathcal{I} = \frac{2 \left\langle \mathcal{F} \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \mathcal{F} \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \mathcal{F} \left(\phi_{P_{2}}\right) \right\rangle} ,
        \end{align*}
        where angled brackets imply averaging.

        F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have an $\phi_{F_{1}}(t) = e^{t}$, in contrast to both parents with $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
